---
title: 'Data Wrangling Assignment #1'
author: "Kenneth Larrieu"
output: html_document
---


``` {r include = FALSE}
library(tidyverse)
```

## Import and Tidy
1. Read the gazetteer data as-is (all columns; no type conversion) into a `gaz_raw` tibble.

``` {r}
gaz_raw = read_delim("CA_Features_20180401.txt", "|", col_types = cols(.default = col_character()))
gaz_raw
```

2. Select columns to include in `gaz` tibble.
``` {r}
gaz = gaz_raw %>% select(FEATURE_ID,
                          FEATURE_NAME,
                          FEATURE_CLASS,
                          STATE_ALPHA,
                          COUNTY_NAME,
                          PRIM_LAT_DEC,
                          PRIM_LONG_DEC,
                          SOURCE_LAT_DEC,
                          SOURCE_LONG_DEC,
                          ELEV_IN_M,
                          MAP_NAME,
                          DATE_CREATED,
                          DATE_EDITED)
gaz
```
3. Convert `gaz` columns to correct type. Convert any placeholders for unknown data to `NA`
```{r}
gaz = gaz %>%
  type_convert(col_types = cols(
    FEATURE_ID = col_integer(),
    PRIM_LAT_DEC = col_double(),
    PRIM_LONG_DEC = col_double(),
    SOURCE_LAT_DEC = col_double(),
    SOURCE_LONG_DEC = col_double(),
    ELEV_IN_M = col_integer(),
    DATE_CREATED = col_date(format = "%M/%d/%Y"),
    DATE_EDITED = col_date(format = "%M/%d/%Y")
          ))

gaz = gaz %>%
  mutate(MAP_NAME = replace(MAP_NAME, MAP_NAME == "Unknown", NA)) %>%
  mutate(PRIM_LAT_DEC = replace(PRIM_LAT_DEC, PRIM_LAT_DEC == 0, NA)) %>%
  mutate(PRIM_LONG_DEC = replace(PRIM_LONG_DEC, PRIM_LONG_DEC == 0, NA)) %>%
  mutate(SOURCE_LAT_DEC = replace(SOURCE_LAT_DEC, SOURCE_LAT_DEC == 0, NA)) %>%
  mutate(SOURCE_LONG_DEC = replace(SOURCE_LONG_DEC, SOURCE_LONG_DEC == 0, NA))
gaz
```


4. Filter out missing primary lat/long, only include places in CA
```{r}
gaz = gaz %>%
  filter(!is.na(PRIM_LAT_DEC) & !is.na(PRIM_LONG_DEC)) %>%
  filter(STATE_ALPHA == "CA")

gaz
```

5. Write out filtered data to CSV
```{r}
write_delim(gaz, "gaz.csv", delim = "|")
```

## Analyze

1. What is the most-frequently-occuring feature name?
```{r}
top_feature = gaz %>%
  group_by(FEATURE_NAME) %>%
  summarize(count = n()) %>%
  filter(count == max(count))

top_feature
```

2. What is the least-frequently-occuring feature class?
```{r}
bottomm_class = gaz %>%
  group_by(FEATURE_CLASS) %>%
  summarize(count = n()) %>%
  filter(count == min(count))
bottomm_class
```

3. What is the approximate center point of each county?
```{r}
#find center of bounding box for each county
county_centers = gaz %>%
  filter(!is.na(COUNTY_NAME)) %>%
  #remove linear features (features with a source lat/long), only want point features
  filter(is.na(SOURCE_LAT_DEC) | is.na(SOURCE_LONG_DEC)) %>%
  group_by(COUNTY_NAME) %>%
  summarize(min_lat = min(PRIM_LAT_DEC), max_lat = max(PRIM_LAT_DEC),
            min_long = min(PRIM_LONG_DEC), max_long = max(PRIM_LONG_DEC)) %>%
  mutate(center_lat = (min_lat+max_lat)/2, center_long = (min_long+max_long)/2) %>%
  select(COUNTY_NAME, center_lat, center_long)
county_centers
```

4. What are the fractions of the total number of features in each county that are natural? man-made?
```{r}
#determine which features are manmade and natural
#some features were ambiguous (could be natural or manmade): Census, Crater, Levee, Park, Trail, Tunnel

class_types = gaz %>%
  group_by(FEATURE_CLASS) %>%
  summarize() %>%
  mutate(type = case_when(FEATURE_CLASS %in% c("Airport",
                                            "Bridge",
                                            "Building",
                                            "Canal",
                                            "Cemetery",
                                            "Church",
                                            "Civil",
                                            "Crossing",
                                            "Dam",
                                            "Harbor",
                                            "Hospital",
                                            "Locale",
                                            "Military",
                                            "Mine",
                                            "Oilfield",
                                            "Populated Place",
                                            "Post Office",
                                            "Reservoir",
                                            "School",
                                            "Tower",
                                            "Well")
                          
                                            ~ "manmade",
                          
                          FEATURE_CLASS %in% c("Arch",
                                               "Area",
                                               "Arroyo",
                                               "Bar",
                                               "Basin",
                                               "Bay",
                                               "Beach",
                                               "Bench",
                                               "Bend",
                                               "Cape",
                                               "Cave",
                                               "Channel",
                                               "Cliff",
                                               "Falls",
                                               "Flat",
                                               "Forest",
                                               "Gap",
                                               "Glacier",
                                               "Gut",
                                               "Island",
                                               "Isthmus",
                                               "Lake",
                                               "Lava",
                                               "Pillar",
                                               "Plain",
                                               "Range",
                                               "Rapids",
                                               "Reserve",
                                               "Ridge",
                                               "Sea",
                                               "Slope",
                                               "Spring",
                                               "Stream",
                                               "Summit",
                                               "Swamp",
                                               "Valley",
                                               "Woods")
                          
                                              ~ "natural"
                          ))

gaz = gaz %>%
  left_join(class_types, by = "FEATURE_CLASS")

#count number of manmade and natural features in each county
types_counts = gaz %>%
  filter(!is.na(COUNTY_NAME)) %>%
  group_by(COUNTY_NAME, type) %>%
  summarize(count = n())

#get fractions
types_counts = types_counts %>%
  mutate(fraction = count/sum(count)) %>%
  select(COUNTY_NAME, type, fraction) %>%
  spread(key = type, value = fraction) %>%
  rename(manmade_fraction = manmade, natural_fraction = natural, ambiguous_fraction = "<NA>")
types_counts
```
